Perturbing SU($n$)-symmetric interactions

### 2020-01-13 collective_sun/sun_perturbation.tex
work out the perturbation theory for more general single- and two-body perturbations

We consider an array of $N$ multilevel spins with non-uniform
SU($n$)-symmetric interactions that can be written in the form
\begin{align}
  H_0 = \sum_{p<q} h_{pq} \Pi_{pq}
  = \f12 \sum_{p,q} h_{pq} \Pi_{pq},
  &&
  \Pi_{pq} \equiv \sum_{\mu,\nu} S_{\mu\nu}^{(p)} S_{\nu\mu}^{(q)},
  \label{eq:ints}
\end{align}
where the coefficients $h_{pq}$ in the Hamiltonian satisfy
$h_{pq}=h_{qp}$ and $h_{pp}=0$;
$S_{\mu\nu}^{(p)}\equiv\op{\mu}{\nu}_p$ flips the state of spin $p$ to
$\ket\mu$ from $\ket\nu$; and the operator $\Pi_{pq}$ permutes spins
$p$ and $q$.  The permutation operator $\Pi_{pq}$ is a multilevel
generalization of the SU(2)-symmetric spin interaction
$\m S_p\c\m S_q$.

We wish to determine the effective dynamics induced on the
ground-state manifold $\M_0$ of $H_0$ by weak single- and two-body
perturbations of the form
\begin{align}
  \V_1 \equiv \sum_p v_p X_p = \v v\c\v X,
  &&
  \V_2 \equiv \sum_{p<q} w_{pq} \times \f12 \p{X_p Y_q + X_q Y_p}
  = \f12 \v X \c\m w \c\v Y,
  \label{eq:perturbations}
\end{align}
where $X,Y$ are self-adjoint\footnote{In future versions of these
  notes, we should relax the restriction that $X,Y$ are self-adjoint.}
trace-zero operators on an $n$-dimensional Hilbert space; $X_p,Y_p$
denote the action of $X,Y$ on spin $p$; $\v X,$ $\v Y$, $\v v$, and
$\m w$ are respectively vectors and a matrix of all
$X_p,Y_p,v_p,w_{pq}$; we use a dot ($\c$) to denote an inner product
between vectors and matrices as appropriate; and we enforce
$w_{pq}=w_{qp}$ and $w_{pp}=0$.
% TODO: consider more general perturbations, namely:
% (1) a sum over X in V_1,
% (2) more general two-body perturbations (O_pq) in V_2, and
% (3) a sum over different O_pq in V_2

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Single-body perturbations}
\label{sec:single_body}

To compute the effective Hamiltonian induced by the single-body
perturbation $\V_1$, we use the permutational symmetry of the
ground-state manifold to simplify
\begin{align}
  H_1^{(1)} = \sum_p v_p \P_0 X_p \P_0
  = \sum_p v_p \P_0 X_0 \P_0 \EQFS \bar v \col{X},
  \label{eq:H_1_1}
\end{align}
where $\EQFS$ denotes equality under a restriction to the fully
symmetric manifold $\M_0$, and
\begin{align}
  \bar v \equiv \f1N \sum_p v_p,
  &&
  \col{X} \equiv \sum_p X_p.
\end{align}
In order to compute the second order effective Hamiltonian
$H_1^{(2)}$, we first choose an arbitrary fully symmetric state
$\ket\psi\in\M_0$ and expand (see Appendix \ref{sec:H_V1_psi}):
\begin{align}
  H_0 \V_1\p{\v v} \ket\psi
  = E_0 \V_1\p{\v v} \ket\psi
  + \p{\sum_{p,q} h_{pq} v_p X_q - \sum_q h_q v_q X_q} \ket\psi,
  &&
  h_q \equiv \sum_p h_{pq}.
  \label{eq:H_V1_psi}
\end{align}
We thus find that $\V_1\p{\v v_\Delta}\ket\psi$ is an eigenvector of
the interaction Hamiltonian $H_0$ if the coefficients $\v v_\Delta$
satisfy
\begin{align}
  \p{\m h - \diag\v h}\c\v v_\Delta = \Delta \v v_\Delta,
  \label{eq:cond_1}
\end{align}
where $\v h,\m h$ are respectively a vector and matrix of all
$h_p,h_{pq}$; $\diag\v h$ is a matrix with $\v h$ on the diagonal and
zeros everywhere else; and $\Delta$ is a scalar.  Solving the general
eigenvalue problem in \eqref{eq:cond_1} requires diagonalizing an
$N\times N$ matrix.  When the interaction Hamiltonian $H_0$ is
translationally invariant, however, this eigenvalue problem can be
solved analytically; we provide the corresponding solution in Appendix
\ref{sec:trans_inv}.

If we {\it construct} a perturbation $\V_1\p{\v v_\Delta}$ with
coefficients $\v v_\Delta$ that satisfy \eqref{eq:cond_1} with
eigenvalue $\Delta$, then
\begin{align}
  H_0 \V_1\p{\v v_\Delta} \ket\psi
  = \p{E_0 + \Delta} \V_1\p{\v v_\Delta} \ket\psi.
\end{align}
Interestingly, the energy of the (unnormalized) state
$\V_1\p{\v v_\Delta}\ket\psi$ depends only on the coefficients
$\v v_\Delta$, and is entirely independent of the state $\ket\psi$ or
the choice of a single-spin trace-zero operator $X$ used to build
$\V_1$ in \eqref{eq:perturbations}.
% TODO: consider more general single-body perturbations V_1, which sum
% over different operators X
Finding operators $\V_1\p{\v v_\Delta}$ that generate eigenvectors of
the interaction Hamiltonian $H_0$ when they act on fully symmetric
states $\ket\psi$ thus reduces to finding eigenvectors $\v v_\Delta$
of the interaction matrix $\m h-\diag\v h$.

If the fully symmetric manifold $\M_0$ is gapped away from all other
states, then then a vector $\V_1\p{\v v_\Delta}\ket\psi$ with
$\Delta=0$ must lie within the fully symmetric manifold $\M_0$, which
implies that the operator $\V_1\p{\v v_\Delta}$ preserves the
permutational symmetry of $\M_0$.  Indeed, a constant vector
$\v\1\equiv\p{1,1,1,\cdots}$ of ones satisfies the condition in
\eqref{eq:cond_1} with eigenvalue $0$.  All other vectors
$\v v_\Delta$ satisfying \eqref{eq:cond_1} with $\Delta\ne0$ must be
orthogonal to $\v\1$, and therefore mean-zero.

We now return to the task of computing the second-order effective
Hamiltonian $H_1^{(2)}$.  Any coefficient vector $\v v$ can be
expanded into its projections $\v v_\Delta$ onto the eigenspace of
vectors satisfying \eqref{eq:cond_1} with eigenvalue $\Delta$,
i.e.~$\v v = \sum_\Delta \v v_\Delta$, which also allows us to expand
\begin{align}
  \V_1\p{\v v} = \sum_\Delta \V_1\p{\v v_\Delta},
\end{align}
where each operator $\V_1\p{\v v_\Delta}$ generates a state with
definite interaction energy $\Delta$ above that of the fully symmetric
manifold $\M_0$.  We can therefore simplify
\begin{align}
  H_1^{(2)}
  = - \P_0 \V_1\p{\v v} \E \V_1\p{\v v} \P_0
  = - \sum_{\Delta\ne0} \f1{\Delta}
  \P_0 \V_1\p{\v v} \P_\Delta \V_1\p{\v v} \P_0
  = - \sum_{\Delta\ne0} \f1{\Delta} \P_0 \V_1\p{\v v_\Delta}^2 \P_0,
\end{align}
where the product $\P_0 \V_1\p{\v v_\Delta}^2 \P_0$ is worked out in
Appendix \ref{sec:PXXP}.  In total, we find
\begin{align}
  H_1^{(2)}
  \EQFS \f1{N\p{N-1}}
  \sum_{\Delta\ne0} \f{\v v_\Delta\c\v v_\Delta}{\Delta}
  \sp{{\col{X}\,}^2 - N \col{X^2}},
  \label{eq:H_1_2}
\end{align}
where $\EQFS$ denotes equality under a restriction to the fully
symmetric manifold $\M_0$, $\v v_\Delta$ is the projection of $\v v$
onto the $\Delta$-eigenspace of the matrix $\m h-\diag\v h$, and
\begin{align}
  \col{X} \equiv \sum_p X_p,
  &&
  \col{X^2} \equiv \sum_p X_p^2.
\end{align}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Two-body perturbations}
\label{sec:two_body}

The first-order effective Hamiltonian $H_2^{(1)}$ induced on the fully
symmetric manifold $\M_0$ by the two-body perturbation $\V_2$ in
\eqref{eq:perturbations} is simply the restriction of $\V_2$ onto
$\M_0$ (see Appendix \ref{sec:PXYP}):
\begin{align}
  H_2^{(1)}
  = \P_0 \V_2 \P_0
  \EQFS \f12 \bar w \p{\col{X}\odot\col{Y} - \col{X\odot Y}},
\end{align}
where $A\odot B\equiv\p{AB+BA}/2$ is a symmetrized product of $A$ and
$B$; and
\begin{align}
  \bar w \equiv {N\choose 2}^{-1} \sum_{p<q} w_{pq},
  &&
  \col{X} \equiv \sum_p X_p,
  &&
  \col{Y} \equiv \sum_p Y_p,
  &&
  \col{X\odot Y} \equiv \sum_p X_p\odot Y_p.
\end{align}
In order to compute the second-order effective Hamiltonian
$H_2^{(2)}$, as before we pick an arbitrary fully symmetric state
$\ket\psi\in\M_0$ and expand (see Appendix \ref{sec:H_V2_psi})
\begin{align}
  H_0 \V_2\p{\m w} \ket\psi
  = E_0 \V_2\p{\m w} \ket\psi
  + \f12 \sum_{p\ne q} \sp{2\bar h_{pq} w_{pq}
    + \sum_k \p{h_{pk} w_{kq} + w_{pk} h_{kq}}}
  X_p Y_q \ket\psi,
  \label{eq:H_V2_psi}
\end{align}
where to keep notation compact, we define
\begin{align}
  \bar h_{pq} \equiv h_{pq} - \f12\p{h_p + h_q},
  &&
  h_p \equiv \sum_q h_{pq}.
\end{align}
We thus find that the vector $\V_2\p{\m w_\Delta}\ket\psi$ is an
eigenvector of the interaction Hamiltonian $H_0$ if the coefficients
$\m w_\Delta$ satisfy the eigenvalue equation
\begin{align}
  \check{\m h} \c \v{\m w}_\Delta = \Delta \v{\m w}_\Delta,
  \label{eq:cond_2}
\end{align}
where for any symmetric $N\times N$ matrix $\m m$ with zeroes on the
diagonal (as is the case with the coefficient matrix $\m w_\Delta$),
we define the ${N \choose 2}$-component vector
\begin{align}
  \v{\m m} \equiv \sum_{p<q} m_{pq} \ket{\set{p,q}},
  \label{eq:flat_vec}
\end{align}
and we define the ${N \choose 2} \times {N \choose 2}$ matrix
\begin{align}
  \check{\m h} = 2\sum_{p<q} \bar h_{pq} \op{\set{p,q}}
  + \sum_{p<q} \ket{\set{p,q}} \sum_{k\notin\set{p,q}}
  \p{h_{pk} \bra{\set{k,q}} + h_{kq} \bra{\set{p,k}}}.
  \label{eq:h_super_mat}
\end{align}
The basis vectors $\ket{\set{p,q}}$ are labeled by a choice
$\set{p,q}$ of two distinct spins $p,q$.  The eigenvalue problem in
\eqref{eq:cond_2} can be simplified when the interaction Hamiltonian
$H_0$ and the perturbation $\V_2\p{\m w}$ are translationally
invariant; we discuss this simplification in Appendix
\ref{sec:trans_inv}.

If we {\it construct} a perturbation $\V_2\p{\m w_\Delta}$ with
coefficients $\m w_\Delta$ that satisfy \eqref{eq:cond_2} with
eigenvalue $\Delta$, then
\begin{align}
  H_0 \V_2\p{\m w_\Delta} \ket\psi
  = \p{E_0 + \Delta} \V_2\p{\m w_\Delta} \ket\psi,
\end{align}
which implies that $\V_2\p{\m w_\Delta}\ket\psi$ is a state of
definite energy that depends only on the coefficients $\m w_\Delta$.
As with the case of single-body perturbations, we can decompose any
matrix $\m w$ into its projections $\m w_\Delta$ onto the eigenspace
of matrices satisfying \eqref{eq:cond_2} with eigenvalue $\Delta$,
i.e.~$\m w=\sum_\Delta\m w_\Delta$, which allows us to expand
\begin{align}
  \V_2\p{\m w} = \sum_\Delta \V_2\p{\m w_\Delta},
\end{align}
and in turn write
\begin{align}
  H_2^{(2)} = - \P_0 \V_2\p{\m w} \E \V_2\p{\m w} \P_0
  = -\sum_{\Delta\ne0} \f1\Delta \P_0 \V_2\p{\m w_\Delta}^2 \P_0,
\end{align}
where the product $\P_0 \V_2\p{\m w_\Delta}^2 \P_0$ is simplified in
Appendices \ref{sec:PXYXYP} and \ref{sec:multi_to_collective}.  If
$X=Y=s_\z$, where $s_\z\equiv\p{\op\up-\op\dn}/2$ is an SU(2) spin
operator, then
\begin{align}
  H_2^{(2)} \EQFS W_4\p{\m h,\m w} S_\z^4 - W_2\p{\m h,\m w} S_\z^2,
\end{align}
where $S_\z\equiv\sum_p s_\z^{(p)}$ is a collective SU(2) spin
operator, $W_k\p{\m h,\m w}$ are scalar coefficients, and we have
neglected scalar terms in $H_2^{(2)}$ without physical consequence.
In order to expand the coefficients $W_k\p{\m h,\m w}$, for any
symmetric matrix $\m m$ with zeros on the diagonal, similarly to the
special definition $\v{\m m}$ in \eqref{eq:flat_vec} we define
\begin{align}
  \v m \equiv \sum_{p,q} m_{pq} \ket{p},
\end{align}
which is essentially the sum over all columns (or rows) of $\m m$, in
terms of which
\begin{align}
  W_4\p{\m h,\m w}
  \equiv \sum_{\Delta\ne0} \f1\Delta \f{\v w_\Delta\c\v w_\Delta
    - \v{\m w}_\Delta\c\v{\m w}_\Delta}{N!_4},
\end{align}
\begin{align}
  W_2\p{\m h,\m w}
  \equiv \sum_{\Delta\ne0} \f1{\Delta} \sp{\f{\v w_\Delta\c\v w_\Delta
      - 2\v{\m w}_\Delta\c\v{\m w}_\Delta}{4N!_2}
    + \f{\p{3N-4}\p{\v w_\Delta\c\v w_\Delta
        - \v{\m w}_\Delta\c\v{\m w}_\Delta}}{2N!_4}},
\end{align}
where
\begin{align}
  N!_k \equiv \prod_{j=0}^{k-1} \p{N-j} = \f{N!}{\p{N-k}!}.
\end{align}
The dependence of the coefficients of $W_k\p{\m h,\m w}$ on the
interaction matrix $\m h$ is implicit in the decomposition of the
coefficient matrix $\m w$ into components $\m w_\Delta$ as
$\m w=\sum_\Delta\m w_\Delta$, where $\m w_\Delta$ is the projection
of $\m w$ onto the $\Delta$-eigenspace of the matrix $\check{\m h}$
defined in \eqref{eq:h_super_mat}.

@@

We wish to simplify
\begin{align}
  H_0 \V_1 \ket\psi
  = \sum_{\substack{k\\p<q}} h_{pq} v_k \Pi_{pq} X_k \ket\psi,
  \label{eq:H_V1_psi_start}
\end{align}
which has terms with $k\in\set{p,q}$, and terms with
$k\notin\set{p,q}$.  In the case of $k\notin\set{p,q}$, the
permutation operator $\Pi_{pq}$ commutes with $X_k$ and annihilates on
$\ket\psi$, leaving us the sum
\begin{align}
  \sum_{\substack{p<q\\k\notin\set{p,q}}} h_{pq}
  = \f12 \sum_{\substack{p,q\\k\notin\set{p,q}}} h_{pq}
  = \f12 \sum_{p,q} h_{pq}
  - \f12 \sum_{\substack{p,q\\k\in\set{p,q}}} h_{pq}
  = E_0 - h_k,
\end{align}
where we have used the facts that $h_{pq}=h_{qp}$ and $h_{pp}=0$.  The
terms in \eqref{eq:H_V1_psi_start} with $k\notin\set{p,q}$ are then
\begin{align}
  \sum_{\substack{p<q\\k\notin\set{p,q}}}
  h_{pq} v_k \Pi_{pq} X_k \ket\psi
  = \sum_k \p{E_0 - h_k} v_k X_k \ket\psi
  = E_0 \V_1 \ket\psi - \sum_k h_k v_k X_k \ket\psi
\end{align}
while the terms with $k\in\set{p,q}$ are
\begin{align}
  \sum_{\substack{p<q\\k\in\set{p,q}}}
  h_{pq} v_k \Pi_{pq} X_k \ket\psi
  = \sum_{k<q} h_{kq} v_k X_q \ket\psi
  + \sum_{p<k} h_{pk} v_k X_p \ket\psi
  = \sum_{k,q} h_{kq} v_k X_q \ket\psi
\end{align}
which implies
\begin{align}
  H_0 \V_1 \ket\psi
  = E_0 \V_1 \ket\psi
  + \p{\sum_{p,q} h_{pq} v_p X_q - \sum_q h_q v_q X_q} \ket\psi.
\end{align}

@@

  \O_S

@@

  N!_{\abs{S}}\, O_S

>>>>

We consider an array of $N$ multilevel spins with non-uniform
SU($n$)-symmetric interactions that can be written in the form
\begin{align}
  H_0 = \sum_{p<q} h_{pq} \Pi_{pq},
  &&
  \Pi_{pq} \equiv \sum_{\mu,\nu} S_{\mu\nu}^{(p)} S_{\nu\mu}^{(q)},
  \label{eq:ints}
\end{align}
where $p,q$ index individual spins; $\mu,\nu$ index states in an
orthonormal basis for the $n$-dimensional Hilbert space $\H_n$ of a
single spin; $h_{pq}$ are scalar coefficients; the operator
$S_{\mu\nu}^{(p)}\equiv\op{\mu}{\nu}_p$ flips the state of spin $p$ to
$\ket\mu$ from $\ket\nu$; and the operator $\Pi_{pq}$ permutes the
states of spins $p$ and $q$.  The permutation operator $\Pi_{pq}$ is a
multilevel generalization of the SU(2)-symmetric spin interaction
$\v S_p\c\v S_q$ for $\v S=\p{S_\x,S_\y,S_\z}$.

We wish to determine the effective dynamics induced on the
ground-state manifold $\M_0$ of $H_0$ by weak single- and two-body
perturbations of the form
\begin{align}
  \V_1 \equiv \sum_X \sum_p v_{Xp} X_p = \sum_X \v v_X\c\v X,
  &&
  \V_2 \equiv \sum_O \sum_{p<q} w_{Opq} O_{pq}
  = \sum_O \v{\m w}_O\c \v{\m O}
  \label{eq:perturbations}
\end{align}
where $p,q$ index individual spins; $v_{Xp}$ and $w_{Opq}$ are real
numbers; $X$ is a trace-zero single-body operator on $\H_n$; $O$ is a
trace-zero two-body operator on $\H_n\otimes\H_n$; $\v v_X$ and $\v X$
are $N$-component vectors defined by
\begin{align}
  \v v_X \equiv \sum_p v_{Xp} \ket{p},
  &&
  \v X \equiv \sum_p X_p \ket{p};
\end{align}
and finally $\v{\m w_O}$ and $\v{\m O}$ are ${N \choose 2}$-component
vectors defined by
\begin{align}
  \v{\m w}_O \equiv \sum_{p<q} w_{Opq} \ket{\set{p,q}},
  &&
  \v{\m O} \equiv \sum_{p<q} O_{pq} \ket{\set{p,q}},
\end{align}
where the basis vectors $\ket{\set{p,q}}$ are labeled by a choice of
two distinct spins $p,q$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Single-body perturbations}
\label{sec:single_body_pert}

To compute the effective Hamiltonian induced by the single-body
perturbation $\V_1$, we use the permutational symmetry of the
ground-state manifold to simplify
\begin{align}
  H_1^{(1)} = \sum_{X,p} v_{Xp} \P_0 X_p \P_0
  \EQFS \sum_{X,p} v_{Xp} X_0
  \EQFS N \sum_X \bar v_X \col{X},
  \label{eq:H_1_1}
\end{align}
where $\EQFS$ denotes equality under a restriction to the fully
symmetric manifold $\M_0$, and
\begin{align}
  \bar v_X \equiv \f1N \sum_p v_{Xp},
  &&
  \col{X} \equiv \f1N \sum_p X_p.
\end{align}
In order to compute the second order effective Hamiltonian
$H_1^{(2)}$, we first choose an arbitrary fully symmetric state
$\ket\psi\in\M_0$ and expand (see Appendix \ref{sec:H_V1_psi}):
\begin{align}
  H_0 \V_1 \ket\psi
  = E_0 \V_1 \ket\psi
  + \sum_X \p{\sum_{p,q} h_{pq} v_{Xp} X_q - \sum_p h_p v_{Xp} X_p}
  \ket\psi,
  &&
  h_p \equiv \sum_q h_{pq},
  \label{eq:H_V1_psi}
\end{align}
where we define $h_{pq}$ for all $p,q$ (as opposed to only $p<q$) by
enforcing $h_{pq}=h_{qp}$ and $h_{pp}=0$.  We thus find that a
perturbation $\V_1^\Delta\ket\psi$ indexed by a scalar $\Delta$ is an
eigenvector of the interaction Hamiltonian $H_0$ if its coefficient
vectors $\v v_X^\Delta$ all satisfy the eigenvalue equation
\begin{align}
  \p{\m h - \diag\v h}\c\v v = \Delta \v v,
  &&
  \m h \equiv \sum_{p,q} h_{pq} \op{p}{q},
  &&
  \v h \equiv \sum_q h_q \ket{q},
  \label{eq:cond_1}
\end{align}
where $\diag\v h\equiv\sum_p h_p \op{p}$ is a matrix with $\v h$ on
the diagonal and zeros everywhere else.  Solving the general
eigenvalue problem in \eqref{eq:cond_1} requires diagonalizing the
$N\times N$ matrix $\m h-\diag\v h$.  When the interaction Hamiltonian
$H_0$ is translationally invariant, however, this eigenvalue problem
can be solved analytically; we provide the corresponding solution in
Appendix \ref{sec:trans_inv}.

If we {\it construct} a perturbation $\V_1^\Delta$ with coefficients
$\v v_X^\Delta$ that satisfy \eqref{eq:cond_1} with eigenvalue
$\Delta$, then
\begin{align}
  H_0 \V_1^\Delta \ket\psi = \p{E_0 + \Delta} \V_1^\Delta \ket\psi.
\end{align}
Interestingly, the energy of the (unnormalized) state
$\V_1^\Delta\ket\psi$ depends only on the coefficients
$\v v_X^\Delta$, and is entirely independent of the state $\ket\psi$
or the choice of a trace-zero single-body operators $X$ used to build
$\V_1^\Delta$.  Finding operators $\V_1^\Delta$ that generate
eigenvectors of the interaction Hamiltonian $H_0$ when they act on
fully symmetric states $\ket\psi$ thus reduces to finding eigenvectors
of the interaction matrix $\m h-\diag\v h$.

If the fully symmetric manifold $\M_0$ is gapped away from all other
states, then then a vector $\V_1^\Delta\ket\psi$ with $\Delta=0$ must
lie within the fully symmetric manifold $\M_0$, which implies that the
operator $\V_1^\Delta$ preserves the permutational symmetry of $\M_0$.
Indeed, a constant vector $\v\1\equiv\p{1,1,1,\cdots}$ of ones
satisfies the condition in \eqref{eq:cond_1} with eigenvalue $0$,
which implies that all vectors $\v v$ satisfying \eqref{eq:cond_1}
with $\Delta\ne0$ must be orthogonal to $\v\1$, and therefore
mean-zero.

We now return to the task of computing the second-order effective
Hamiltonian $H_1^{(2)}$.  Any coefficient vector $\v v_X$ can be
expanded into its projections $\v v_X^\Delta$ onto the eigenspace of
vectors satisfying \eqref{eq:cond_1} with eigenvalue $\Delta$,
i.e.~$\v v_X = \sum_\Delta \v v_X^\Delta$, which also allows us to
expand
\begin{align}
  \V_1 = \sum_\Delta \V_1^\Delta,
  &&
  \V_1^\Delta \equiv \sum_X \v v_X^\Delta \c \v X,
\end{align}
where each operator $\V_1^\Delta$ generates a state with definite
interaction energy $\Delta$ above that of the fully symmetric manifold
$\M_0$.  We can therefore simplify
\begin{align}
  H_1^{(2)}
  = - \P_0 \V_1 \E \V_1 \P_0
  = - \sum_{\Delta\ne0} \f1{\Delta}
  \P_0 \V_1^\Delta \P_\Delta \V_1^\Delta \P_0
  = - \sum_{\Delta\ne0} \f1{\Delta} \P_0 \p{\V_1^\Delta}^2 \P_0,
\end{align}
where the product $\P_0 \p{\V_1^\Delta}^2 \P_0$ is worked out in
Appendix \ref{sec:PXYP}.  In total, we find
\begin{align}
  H_1^{(2)}
  \EQFS \f{N}{N-1} \sum_{X,Y} \sum_{\Delta\ne0}
  \f{\v v_X^\Delta\c\v v_Y^\Delta}{\Delta}
  \p{\col{X}\,\col{Y} - \col{XY}},
  \label{eq:H_1_2}
\end{align}
where $\EQFS$ denotes equality under a restriction to the fully
symmetric manifold $\M_0$; $\v v_X^\Delta$ is the projection of
$\v v_X$ onto the $\Delta$-eigenspace of the matrix $\m h-\diag\v h$;
and for any single-body operator $Z$, we define the spin-averaged
operator
\begin{align}
  \col{Z} \equiv \f1N \sum_p Z_p.
\end{align}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Two-body perturbations}
\label{sec:two_body_pert}

The first-order effective Hamiltonian $H_2^{(1)}$ induced on the fully
symmetric manifold $\M_0$ by the two-body perturbation $\V_2$ in
\eqref{eq:perturbations} is simply the restriction of $\V_2$ onto
$\M_0$:
\begin{align}
  H_2^{(1)} = \sum_O \sum_{p<q} w_{Opq} \P_0 O_{pq} \P_0
  \EQFS \sum_O \sum_{p<q} w_{Opq} O_{0,1}
  \EQFS {N \choose 2} \sum_O \bar w_O \col{O},
\end{align}
where
\begin{align}
  \bar w_O \equiv {N \choose 2}^{-1} \sum_{p<q} w_{Opq},
  &&
  \col{O} \equiv {N \choose 2}^{-1} \sum_{p<q} O_{pq},
\end{align}
are averages of $w_{Opq}$ and $O_{pq}$ over all pairs of spins $p,q$.
If the two-body operator $O=X\otimes Y$, then
\begin{align}
  \col{O} \EQFS X_0 Y_1
  \EQFS \f1{N-1} \p{N\, \col{X}\,\col{Y} - \col{XY}},
\end{align}
so
\begin{align}
  H_2^{(1)} \EQFS \f12 N \sum_{\p{X,Y}} \bar w_{\p{X,Y}}
  \p{N\, \col{X}\,\col{Y} - \col{XY}}.
\end{align}
In order to compute the second-order effective Hamiltonian
$H_2^{(2)}$, as before we pick an arbitrary fully symmetric state
$\ket\psi\in\M_0$ and expand (see Appendix \ref{sec:H_V2_psi})
\begin{align}
  H_0 \V_2 \ket\psi
  = E_0 \V_2 \ket\psi
  + \f12 \sum_{p\ne q} \sp{2\bar h_{pq} w_{Opq}
    + \sum_k \p{h_{pk} w_{Okq} + w_{Opk} h_{kq}}}
  O_{pq} \ket\psi,
  \label{eq:H_V2_psi}
\end{align}
where to keep notation compact, we define
\begin{align}
  \bar h_{pq} \equiv h_{pq} - \f12\p{h_p + h_q},
  &&
  h_p \equiv \sum_q h_{pq},
\end{align}
and we define $w_{Opq}$ for all $p,q$ (as opposed to only $p<q$) by
enforcing $w_{Opq}=w_{Oqp}$ and $w_{Opp}=0$.  We thus find that the
vector $\V_2^\Delta\ket\psi$ indexed by a scalar $\Delta$ is an
eigenvector of the interaction Hamiltonian $H_0$ if its coefficients
$\v{\m w}_O^\Delta$ all satisfy the eigenvalue equation
\begin{align}
  \check{\m h} \c \v{\m w} = \Delta \v{\m w},
  \label{eq:cond_2}
\end{align}
where
\begin{align}
  \check{\m h} \equiv 2\sum_{p<q} \bar h_{pq} \op{\set{p,q}}
  + \sum_{p<q} \ket{\set{p,q}} \sum_{k\notin\set{p,q}}
  \p{h_{pk} \bra{\set{k,q}} + h_{kq} \bra{\set{p,k}}}.
  \label{eq:h_super_mat}
\end{align}
The eigenvalue problem in \eqref{eq:cond_2} can be simplified somewhat
when the interaction Hamiltonian $H_0$ is translationally invariant;
we discuss this simplification in Appendix \ref{sec:trans_inv}.

If we {\it construct} a perturbation $\V_2^\Delta$ with coefficients
$\v{\m w}_O^\Delta$ that satisfy \eqref{eq:cond_2} with eigenvalue
$\Delta$, then
\begin{align}
  H_0 \V_2^\Delta \ket\psi = \p{E_0 + \Delta} \V_2^\Delta \ket\psi,
\end{align}
which implies that $\V_2^\Delta\ket\psi$ is an (unnormalized0 state of
definite energy that depends only on the coefficients $\m w_O^\Delta$.
Similarly to the case of single-body perturbations, we can decompose
any vector $\v{\m w}_O$ into its projections $\v{\m w}_O^\Delta$ onto
the eigenspace of vectors satisfying \eqref{eq:cond_2} with eigenvalue
$\Delta$, i.e.~$\v{\m w}_O=\sum_\Delta\v{\m w}_O^\Delta$, which allows
us to expand
\begin{align}
  \V_2 = \sum_\Delta \V_2^\Delta,
  &&
  \V_2^\Delta \equiv \sum_O \v{\m w}_O^\Delta \c \v{\m O},
\end{align}
and in turn write
\begin{align}
  H_2^{(2)} = - \P_0 \V_2^\Delta \E \V_2^\Delta \P_0
  = -\sum_{\Delta\ne0} \f1\Delta \P_0 \p{\V_2^\Delta}^2 \P_0,
\end{align}
where the product $\P_0 \p{\V_2^\Delta}^2 \P_0$ is simplified in
Appendices \ref{sec:POQP} and \ref{sec:multi_to_collective}.  In the
special case that
\begin{align}
  \V_2 = \sum_{p<q} w_{pq} s_\z^{(p)} s_\z^{(q)},
\end{align}
where $s_\z\equiv\p{\op\up-\op\dn}/2$ is an SU(2) spin-$z$ operator,
then
\begin{align}
  H_2^{(2)} \EQFS W_4 S_\z^4 - W_2 S_\z^2,
\end{align}
where $S_\z\equiv\sum_p s_\z^{(p)}$ is a collective SU(2) spin
operator, $W_k$ are scalar coefficients, and we have neglected scalar
terms in $H_2^{(2)}$ without physical consequence.  In order to expand
the coefficients $W_k$, for any ${N \choose 2}$-component vector
$\v{\m m}=\sum_{p<q}m_{pq}\ket{\set{p,q}}$ we define the $N$-component
vector
\begin{align}
  \v m \equiv \sum_{p,q} m_{pq} \ket{p},
\end{align}
where we define $m_{pq}$ for all $p,q$ (as opposed to only $p<q$) by
enforcing $m_{pq}=m_{qp}$ and $m_{pp}=0$.  We then denote the
projection of $\v{\m m}$ onto the $\Delta$-eigenspace of the matrix
$\check{\m h}$ defined in \eqref{eq:h_super_mat} by $\v{\m m}_\Delta$,
which allows us to expand
\begin{align}
  W_4
  \equiv \sum_{\Delta\ne0} \f1\Delta \f{\v w_\Delta\c\v w_\Delta
    - \v{\m w}_\Delta\c\v{\m w}_\Delta}{N!_4},
\end{align}
and
\begin{align}
  W_2
  \equiv \sum_{\Delta\ne0} \f1{\Delta} \sp{\f{\v w_\Delta\c\v w_\Delta
      - 2\v{\m w}_\Delta\c\v{\m w}_\Delta}{4N!_2}
    + \f{\p{3N-4}\p{\v w_\Delta\c\v w_\Delta
        - \v{\m w}_\Delta\c\v{\m w}_\Delta}}{2N!_4}},
\end{align}
where
\begin{align}
  N!_k \equiv \prod_{j=0}^{k-1} \p{N-j} = \f{N!}{\p{N-k}!}.
\end{align}

@@

We wish to simplify
\begin{align}
  H_0 \V_1 \ket\psi
  = \sum_{X,k} \sum_{p<q} h_{pq} v_{Xk} \Pi_{pq} X_k \ket\psi,
  \label{eq:H_V1_psi_start}
\end{align}
which has terms with $k\in\set{p,q}$, and terms with
$k\notin\set{p,q}$.  In the case of $k\notin\set{p,q}$, the
permutation operator $\Pi_{pq}$ commutes with $X_k$ and annihilates on
$\ket\psi$, leaving terms at each fixed $k$ of the form
\begin{align}
  \sum_{\substack{p<q\\k\notin\set{p,q}}} h_{pq}
  = \sum_{p<q} h_{pq} - \sum_{\substack{p<q\\k\in\set{p,q}}} h_{pq}
  = E_0 - \f12 \sum_{\substack{p,q\\k\in\set{p,q}}} h_{pq}
  = E_0 - h_k,
\end{align}
where we define $h_{pq}$ for all $p,q$ (as opposed to only $p<q$) by
enforcing $h_{pq}=h_{qp}$ and $h_{pp}=0$; and
$h_k \equiv \sum_\ell h_{k\ell}$.  The terms in
\eqref{eq:H_V1_psi_start} with $k\notin\set{p,q}$ are then
\begin{align}
  \sum_X \sum_{\substack{p<q\\k\notin\set{p,q}}}
  h_{pq} v_{Xk} \Pi_{pq} X_k \ket\psi
  = \sum_{X,k} \p{E_0 - h_k} v_{Xk} X_k \ket\psi
  = E_0 \V_1 \ket\psi - \sum_{X,k} h_k v_{Xk} X_k \ket\psi
\end{align}
while the terms in \eqref{eq:H_V1_psi_start} with $k\in\set{p,q}$ for
each fixed $X$ take the form
\begin{align}
  \sum_{\substack{p<q\\k\in\set{p,q}}}
  h_{pq} v_{Xk} \Pi_{pq} X_k \ket\psi
  = \sum_{k<q} h_{kq} v_{Xk} X_q \ket\psi
  + \sum_{p<k} h_{pk} v_{Xk} X_p \ket\psi
  = \sum_{k,q} h_{kq} v_{Xk} X_q \ket\psi
\end{align}
which implies
\begin{align}
  H_0 \V_1 \ket\psi
  = E_0 \V_1 \ket\psi
  + \sum_X \p{\sum_{p,q} h_{pq} v_{Xp} X_q - \sum_p h_p v_{Xp} X_p}
  \ket\psi.
\end{align}

@@

  N^{\abs{S}} \O_S

@@

  N!_{\abs{S}} O_S
